Mergeable Types

### 2017-02-28 paper/evaluation.tex
Semantics.
edits

In this section, we present an evaluation of \name programming model with the
aim of assessing its practical utility for describing geo-distributed
eventually consistent applications. In particular, we are interested in the
horizontal scalability of applications built using \name. Typical distributed
eventually consistent data stores use custom synchronization and dissemination
protocols for transferring updates between replicas. Unlike this, \name is
realized over Irmin which uses Git transfer protocol~\cite{..}, a protocol not designed
with distributed data stores in mind. Hence, we also evaluate the performance
of synchronization and merging.

@@

In order to evaluate this, we have implemented a collaborative editing
application that simulates concurrent editing of the same document by several
authors. This application is not unlike Google Docs, but also differs from it
by the fact that we do not need a central server that coordinates the edits.
Instead, edits from each author is asynchronously sent to other authors. The
shared document itself is represented by a mergeable rope of characters
(Section~\ref{sec:model}), and hence, the remote edits can be correctly
reconciled into the local document. We resolve conflicts on concurrent
substitutions at the same position (such as two authors concurrently editing
\C{"abc"} to \C{"xbc"} and \C{"ybc"}, respectively) by choosing the character
with the smaller ASCII code. For the benchmarks, the initial document that we
use has 1576 characters. The workload consists of 4000 edit operations at
random indices with 85\% insertions and 15\% deletions.

@@

\begin{figure}[t]
  \centering
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/cluster.pdf}
		\caption{8-node ring cluster on Google Cloud Platform used for
			benchmarking. US-C, EU-W, ASIA-NE and ASIA-E are availability zones
			us-central1-b, europe-west1-b, asia-northeast1-b and asia-east1-b
			respectively. Edge labels are inter-node latencies in milliseconds.}
		\label{fig:cluster}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{Graphs/scalability.pdf}
		\caption{Scalability: Overall throughput of the cluster and latency of each
			operation.}
		\label{grf:scalability}
	\end{subfigure}
	\par\bigskip
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{Graphs/synchronization.pdf}
		\caption{Synchronization: Overall throughput of the cluster while actively
			synchronizing with the successor node.}
		\label{grf:synchronization}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{Graphs/merge.pdf}
		\caption{Merge performance: Cost of merging concurrent operations across
		nodes.}
		\label{grf:merge}
	\end{subfigure}
	\caption{\name performance evaluation.}
  \label{grf:evaluation}
\end{figure}

For our experiments, we instantiate an 8-node geo-distributed cluster in Google
Cloud Platform~\cite{gcp}. Each node is an \C{n1-standard-4} instance with 2
virtual CPUs and 7.5 GB of memory. The nodes are arranged in a ring as shown in
Figure~\ref{fig:cluster} and synchronize (push and pull updates) when necessary
with its successor. While \name does not impose any restrictions on
synchronizing with other nodes, we choose to only synchronize with the
successor since we are guaranteed eventual convergence with a ring cluster. If
the successor becomes unavailable, the node synchronizes with other available
nodes. Operations are generated and applied locally on each node before they
are propagated eventually to other nodes with the help of a background thread
that synchronizes with its successor every second.

@@

Finally, we evaluate the performance of merge functions. For this experiment,
we consider the nodes 1 (\C{us-central1-b}) and 2 (\C{europe-west1-b}). Both
nodes start with the initial document and perform \C{n} concurrent edits
without synchronization. After this, the nodes synchronize pushing the local
edits and pulling the remote edits, and merge to produce the final document. In
this case, we indicate the distance of synchronization as \C{2n}.
Figure~\ref{grf:merge} presents the time share between network and computation
for performing 3-way merge as we increase the operation distance between the
remote branches. We see that the merge time is dominated by the network
overhead. This is because of the transfer protocol involving multiple
round trips and the need to transfer objects not only for application objects
but also Git internal objects for history, directory structure and branches. We
observe that we can improve the network performance by implementing a streaming
protocol that eagerly transfers objects.

More importantly, the cost of computation grows sub-linearly. This is because
of efficient merge function on ropes~\ref{ropes} enabled by Irmin's
content-addressed storage backend, and more expensive operational
transformation of merging concurrent edits to leaf nodes~\ref{lists} is invoked
rarely on small. On average, we measured that at a distance of 256, for merging
the two documents with 1743 characters, 50 list merges were performed on
strings of average length 13. A purely list based approach would require the
polynomial time edit sequence algorithm to be applied on the entire document
twice. Compare to this, the rope implementation has much better algorithmic
complexity. We measured that by representing the entire document using
mergeable list type, in addition to slowing down each edit operation, the merge
at a distance of 256 was 2.3 times slower than the rope implementation.

>>>>

In this section, we present an evaluation of \name programming model
with the aim of assessing its practical utility for describing
geo-distributed eventually consistent applications. In particular, we
are interested in the horizontal scalability of applications built
using \name. Typical distributed eventually consistent data stores use
custom synchronization and dissemination protocols for transferring
updates between replicas. In contrast, \name is realized over Irmin~\cite{Irmin}
which uses Git transfer protocol~\cite{..}, a protocol not designed
with distributed data stores in mind. Hence, we also evaluate the
performance of synchronization and merging.

@@

To evaluate the effectiveness of our model, we have implemented a
collaborative editing application that simulates concurrent editing of
the same document by several authors. This application is not unlike
Google Docs, but differs from it by the fact that we do not need a
central server that coordinates the edits.  Instead, edits from each
author is asynchronously sent to other authors. The shared document
itself is represented by a mergeable rope of characters
(Section~\ref{sec:model}), and hence, the remote edits can be
correctly reconciled into each local document. We resolve conflicts on
concurrent substitutions at the same position (such as two authors
concurrently editing \C{"abc"} to \C{"xbc"} and \C{"ybc"},
respectively) by choosing the character with the smaller ASCII
code. For the benchmarks, the initial document that we use has 1576
characters. The workload consists of 4000 edit operations at random
indices with 85\% insertions and 15\% deletions.

@@

\begin{figure}[t]
  \centering
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/cluster.pdf}
		\caption{Our experimental configuration consists of an
                  8-node ring cluster on executing on a Google Cloud
                  Platform. US-C, EU-W, ASIA-NE and ASIA-E are
                  availability zones us-central1-b, europe-west1-b,
                  asia-northeast1-b and asia-east1-b
                  respectively. Edge labels are inter-node latencies
                  in milliseconds.}
		\label{fig:cluster}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{Graphs/scalability.pdf}
		\caption{Scalability: Overall throughput of the cluster and latency of each
			operation.}
		\label{grf:scalability}
	\end{subfigure}
	\par\bigskip
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{Graphs/synchronization.pdf}
		\caption{Synchronization: Overall throughput of the cluster while actively
			synchronizing with the successor node.}
		\label{grf:synchronization}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth]{Graphs/merge.pdf}
		\caption{Merge performance: Cost of merging concurrent operations across
		nodes.}
		\label{grf:merge}
	\end{subfigure}
	\caption{\name performance evaluation.}
  \label{grf:evaluation}
\end{figure}

For our experiments, we instantiate an 8-node geo-distributed cluster
using the Google Cloud Platform~\cite{gcp}. Each node is an
\C{n1-standard-4} instance with 2 virtual CPUs and 7.5 GB of
memory. The nodes are arranged in a ring as shown in
Figure~\ref{fig:cluster} and synchronize (push and pull updates) when
necessary with its successor. While \name does not impose any
restrictions on synchronizing with other nodes, we choose to only
synchronize with the successor since we are guaranteed eventual
convergence with a ring cluster. If the successor becomes unavailable,
the node synchronizes with other available nodes. Operations are
generated and applied locally on each node before they are propagated
eventually to other nodes with the help of a background thread that
synchronizes with its successor every second.

@@

Finally, we evaluate the performance of merge functions. For this experiment,
we consider the nodes 1 (\C{us-central1-b}) and 2 (\C{europe-west1-b}). Both
nodes start with the initial document and perform $n$ concurrent edits
without synchronization. After this, the nodes synchronize pushing the local
edits and pulling the remote edits, and merge to produce the final document. In
this case, we indicate the distance of synchronization as $2n$.
Figure~\ref{grf:merge} presents the time share between network and computation
for performing 3-way merge as we increase the operation distance between the
remote branches. We see that the merge time is dominated by the network
overhead. This is because of the transfer protocol involving multiple
round trips and the need to transfer objects not only for application objects
but also Git internal objects for history, directory structure and branches. We
observe that we can improve the network performance by implementing a streaming
protocol that eagerly transfers objects.

More importantly, the cost of computation grows sub-linearly. This is
because of efficient merge function on ropes (Sec.~\ref{sec:ropes})
enabled by Irmin's content-addressed storage backend, and more
expensive operational transformation of merging concurrent edits to
leaf nodes (Sec.~\ref{sec:lists}) is invoked rarely on small. On
average, we measured that with a distance of 256, for merging the two
documents with 1743 characters, 50 list merges were performed on
strings of average length 13. A purely list based approach would
require a polynomial time edit sequence algorithm to be applied on the
entire document twice. Compared to this, the rope implementation has
much better algorithmic complexity.  While not depicted here, we
measured that by representing the entire document using a mergeable
list type, in addition to slowing down each edit operation, the merge
at a distance of 256 was 2.3 times slower than the rope
implementation.  We conclude that versioning and merging as found in
\name\ can yield substantial throughput and latency gains with low
computational cost on modern eventually-consistent cloud systems.

### 2017-02-28 paper/examples.tex
edits

\noindent The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The implementation
		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

@@

A mergeable list of integers can be composed using mergeable lists and
mergeable integers.  Concretely, it is a composition of \C{MList} and
\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
as \C{Counter.merge} (Sec.~\ref{sec:intro}).  An alternative
instantiation of \C{MList} is for a quantity type
(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
\begin{wrapfigure}{L}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
  \vspace*{-.2in}
\end{wrapfigure}
defines its \C{merge} operation a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

@@

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is shopping list application, that allows its users to
collaboratively build a shopping list (listing items in the decreasing
order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
\begin{wrapfigure}{L}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{-.2in}
\end{wrapfigure}
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
to justify using ropes, but \C{MRope(MList(Item))} is definitely a
possibility).  \C{MList(Item).merge} automatically lifts the merge
semantics for items to a merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
and Bob's quantity as per the definition of \C{Qty.merge}.

The examples described above make certain assumptions about the
concurrency model and the underlying system, such as the existence of
a single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.

>>>>

\noindent The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The implementation
		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (shown in the Appendix) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

@@

A mergeable list of integers can be composed using mergeable lists and
mergeable integers.  Concretely, it is a composition of \C{MList} and
\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
as \C{Counter.merge} described in the introduction.  An alternative
instantiation of \C{MList} is for a quantity type
(\C{Qty}) that supports \C{add} and \C{subtract} operations, and
defines its \C{merge} operation a little differently:
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
  \vspace*{-.45in}
\end{wrapfigure}
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is a shopping list application that allows its users
to collaboratively build a shopping list (listing items in the
decreasing order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}).  The
merge function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{-.35in}
\end{wrapfigure}
i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
to justify using ropes, but \C{MRope(MList(Item))} is certainly
allowed).  \C{MList(Item).merge} automatically lifts the merge
semantics for items to a merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
and Bob's quantity as per the definition of \C{Qty.merge}.
The examples described above make certain assumptions about the
concurrency model and the underlying system, such as the existence of
a single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.

@@
